\pdftitle

### 2018-05-18 maxNt_v3.tex
V3

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-18T10:28:29+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

The purpose of the present work is not to assess the merits of \me\ methods
with respect to other methods. Its main purpose is to show that there is a
problem in the way the \me\ method itself, as sketched above, is applied to
the activity of the recorded neurons. We believe that this problem is at
the root of some quirks about this method that were pointed out in the
literature \cite{roudietal2009b}. This problems extends also to more
complex versions of the method, possibly except version that use
\enquote{hidden} neurons. The problem is that the recorded neurons are a
\emph{sample} from a larger, unrecorded population, but the \me\ method as
applied above is treating them as isolated from the rest of the brain.
Hence, the results it provides cannot rightfully be extrapolated. We will
give a mathematical proof of this. Let us first analyse this issue in more
detail.

@@

The latter formula, in particular, shows that the standard way of applying
\me\ (without \enquote{hidden neurons}) implicitly assumes that \emph{no}
larger population exists beyond the recorded sample of neurons. One could
in fact object to the application at the population level, and say that the
traditional way of applying \me, \eqn~\eqref{eq:app_maxent_sample}, yields
different results because it does not make assumptions about the size
$\yNv$ of a possibly existing larger population. Such a state of
uncertainty, however, is correctly formalized according to the laws of
probability by introducing a probability distribution for $\yNv$, and is
expressed by \eqn~\eqref{eq:app_maxent_pop_marg_unknown_N}. This expression
cannot generally be equal to~\eqref{eq:app_maxent_sample} unless the
distribution for $\yNv$ gives unit probability to $\yNv=n$; that is,
unless the sample \emph{is} the full population, and no larger population
exists or the recorded neurons are completely isolated from it.

The latter notion is unreasonable. It amounts to say that the neurons we
distinguished and tracked with our recording device are a very special,
isolated set among all those that could have been recorded. It is for this
reason that we find the \me\ application at the population level
preferable. Physical models of neuronal networks usually include some sort
of external input, mimicking an embedding in a larger network. The \me\
distribution obtained at the population level may reveal features in a
data set that were unnoticed by the standard \me\ model.

The difference between the applications at the sample and at the population
levels appears also in statistical mechanics. If a system is statistically
described by a \me\ Gibbs state, its subsystems cannot be described by a
Gibbs state \cite{maesetal1999}. A somewhat similar situation also appears
in the statistical description of the final state of a non-equilibrium
process starting and ending in two equilibrium states: we can describe our
knowledge about the final state either by (1) a Gibbs distribution,
calculated from the final equilibrium macrovariables, or (2) by the
distribution obtained from the Liouville evolution of the Gibbs
distribution assigned to the initial state. The two distributions differ
(even though the final \emph{physical} state is obviously exactly the same
\cite[\sect~4]{jaynes1985d_r1993}), and the second allows us to make
sharper predictions about the final physical state thanks to our knowledge
of its preceding dynamics. In this example, though, both distributions are
usually immensely sharp and practically lead to the same predictions. In
the neuroscientific applications considered in this note the difference in
predictions of the sample \vs\ full-population applications may be very
relevant instead.

The idea of the new application leads in fact to more questions. For
instance:
\begin{itemize}
\item Do the standard and new applications lead to different or contrasting
  conclusions about \enquote{cooperativity}, when applied to real data
  sets?
\item How to extend the new application to the \enquote{inhomogeneous} case
  \cite{schneidmanetal2006,shlensetal2006,roudietal2009b}, in which
  expectations for individual neurons or groups of neurons are constrained?
\item What is the mathematical relation between the new application and
  \me\ models with hidden neurons?
\end{itemize}
Owing to space limitations we cannot investigate these questions here, but
we hope to in future work.

@@

\newcommand*{\citein}[2][]{\textnormal{\textcite[#1]{#2}}%\addtocategory{extras}{#2}
}
\newcommand*{\citebi}[2][]{ref.\ \citep[#1]{#2}%\addtocategory{extras}{#2}
}
\newcommand*{\subtitleproc}[1]{}

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-18T11:33:51+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

The purpose of the present work is not to assess the merits of \me\ methods
with respect to other methods. Its main purpose is to show that there is a
problem in the way the \me\ method itself, as sketched above, is applied to
the activity of the recorded neurons. We believe that this problem is at
the root of some quirks about this method that were pointed out in the
literature \cite{roudietal2009b}. This problems extends also to more
complex versions of the method, possibly except version that use
\enquote{hidden} neurons \cite{smolensky1986,hintonetal2006}. The problem
is that the recorded neurons are a \emph{sample} from a larger, unrecorded
population, but the \me\ method as applied above is treating them as
isolated from the rest of the brain. Hence, the results it provides cannot
rightfully be extrapolated. We will give a mathematical proof of this. Let
us first analyse this issue in more detail.

@@

The latter formula, in particular, shows that the standard way of applying
\me\ 
implicitly assumes that \emph{no} larger population exists beyond the
recorded sample of neurons. One could in fact object to the application at
the population level, and say that the traditional way of applying \me,
\eqn~\eqref{eq:app_maxent_sample}, yields different results because it does
not make assumptions about the size $\yNv$ of a possibly existing larger
population. Such a state of uncertainty, however, is correctly formalized
according to the laws of probability by introducing a probability
distribution for $\yNv$, and is expressed by
\eqn~\eqref{eq:app_maxent_pop_marg_unknown_N}. This expression cannot
generally be equal to~\eqref{eq:app_maxent_sample} unless the distribution
for $\yNv$ gives unit probability to $\yNv=n$; that is, unless the sample
\emph{is} the full population, and no larger population exists or the
recorded neurons are completely isolated from it.

This hidden assumption of isolation is unreasonable. It amounts to say that
the neurons we distinguished and tracked with our recording device are a
very special, isolated set among all those that could have been recorded.
It is for this reason that we find the \me\ application at the population
level preferable. Physical models of neuronal networks usually include some
sort of external input, mimicking an embedding in a larger network. The
\me\ distribution obtained at the population level may reveal features in a
data set that were unnoticed by the standard \me\ model.

The difference in the resulting distributions between the applications at
the sample and at the population levels appears in the use of Boltzmann
machines with hidden units \cite{lerouxetal2008}, although by a different
conceptual route. It also appears in statistical mechanics: if a system is
statistically described by a \me\ Gibbs state, its subsystems cannot be
described by a Gibbs state \cite{maesetal1999}. A somewhat similar
situation also appears in the statistical description of the final state of
a non-equilibrium process starting and ending in two equilibrium states: we
can describe our knowledge about the final state either by (1) a Gibbs
distribution, calculated from the final equilibrium macrovariables, or (2)
by the distribution obtained from the Liouville evolution of the Gibbs
distribution assigned to the initial state. The two distributions differ
(even though the final \emph{physical} state is obviously exactly the same
\cite[\sect~4]{jaynes1985d_r1993}), and the second allows us to make
sharper predictions about the final physical state thanks to our knowledge
of its preceding dynamics. In this example, though, both distributions are
usually immensely sharp and practically lead to the same predictions. In
neuroscientific applications, the difference in predictions of the sample
\vs\ full-population applications can instead be very relevant.

The idea of the new application leads in fact to more questions. For
instance:
\begin{itemize}
\item Do the standard and new applications lead to different or contrasting
  conclusions about \enquote{cooperativity}, when applied to real data
  sets?
\item How to extend the new application to the \enquote{inhomogeneous} case
  \cite{schneidmanetal2006,shlensetal2006,roudietal2009b}, in which
  expectations for individual neurons or groups of neurons are constrained?
\item What is the mathematical relation between the new application and
  \me\ models with hidden neurons \cite{smolensky1986,hintonetal2006}?
\end{itemize}
Owing to space limitations we cannot investigate these questions here, but
we hope to in future work.

@@

\newcommand*{\citein}[2][]{\textnormal{\textcite[#1]{#2}}%\addtocategory{extras}{#2}
}
\newcommand*{\citebi}[2][]{ref.\ \citep[#1]{#2}%\addtocategory{extras}{#2}
}
\newcommand*{\subtitleproc}[1]{}
\newcommand*{\chapb}{ch.}
